---
title: "Hungry Guys Radio Spotify Analysis"
author: "Nik Linnane"
output: html_notebook
---

# Import Libraries & Data
```{r}
library(pacman)
options(scipen=999)
p_load(tidyverse, lubridate, dplyr, ggplot2, reshape, magick, ggridges, ggExtra, treemap, fmsb)
data <- read_csv('/Users/nik.linnane/Downloads/Spotify/HungryGuysRadio.csv')
logo2 <- image_read('/Users/nik.linnane/Downloads/Spotify/hungryguyslogo2.png')
logo3 <- image_read('/Users/nik.linnane/Downloads/Spotify/spotlogo.png')
# standardize columns names
names(data) <- toupper(names(data))
```

# Exploratory Data Analysis
```{r}
# create long dataframe of audio features
audioFeaturesLong <- data %>% 
  select(DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, ACOUSTICNESS, POPULARITY, TEMPO, DURATION_MIN) %>% 
  pivot_longer(DANCEABILITY:DURATION_MIN, names_to='Audio Feature', values_to='Value')
```

```{r}
# what does the distribution of audio features look like across the playlist?
audioFeaturesLong %>% 
  ggplot(aes(Value)) +
  facet_wrap(~`Audio Feature`, scales = 'free', ncol = 4, strip.position = 'top') +
  geom_histogram(binwidth = function(x) 2 * IQR(x) / (length(x)^(1/3)), fill='#1cd155') +
  labs(title = 'Audio Feature Distributions',
       subtitle = 'Hungry Guys Radio on Spotify',
       caption = 'Check out and follow the playlist by scanning this Spotify code!
       Data acquired via "Spotipy" API',
       x = '',
       y = '') +
  theme_grey() +
  theme(plot.title = element_text(color = 'black', face = 'bold', size = 16, hjust = 0.5),
        plot.caption = element_text(color = "black", face = 'italic', hjust = 1),
        plot.subtitle = element_text(color = "black", face = 'italic', hjust = 0.5),
        strip.text = element_text(size=8))
  grid::grid.raster(logo3, 
                    x = 0.93, 
                    y = 0.9, 
                    just = c('left', 'bottom'), 
                    width = unit(0.35, 'inches'))
  grid::grid.raster(logo2, 
                    x = 0.79, 
                    y = 0.13, 
                    just = c('left', 'bottom'), 
                    width = unit(1.5, 'inches'))

```

initial audio features insights...
- more non-acoustic songs
- consists of a plurality of mostly dancable songs
- a couple of songs at or longer than 6 minutes
- more high-energy
- next to none songs that resemble an instrumental or a live recording
- songs tend to be more loud
- there are some mildly popular songs, however, there are a large proportion of songs that are very unpopular
- most songs consist of music and words
- most songs hover around 100 bpm
- decent distribution of valence but definitely more happy than not
- how does all of this compare to other types of playlists? Top 100, different moods and genres etc.

# Loudness
```{r}
# who has most contributed to the loudness of the playlist?
data$USER <- factor(data$USER, levels = c('Ian', 'Conor', 'Jack', 'Nik'))
data %>% 
  ggplot(aes(x=LOUDNESS, y=USER, fill=USER)) +
  geom_density_ridges(alpha=0.8) +
  labs(title = 'Loudness Contribution by User',
       subtitle = 'Hungry Guys Radio on Spotify',
       caption = 'Nik contributes most to the loudness of the playlist, centered around -5db.
    Data acquired via "Spotipy" API',
       x = '',
       y = '') +
  theme_grey() + 
  theme(legend.position = 'none',
        plot.title = element_text(color = 'black', face = 'bold', size = 16, hjust = 0.5),
        plot.caption = element_text(color = "black", face = 'italic', hjust = 0),
        plot.subtitle = element_text(color = "black", face = 'italic', hjust = 0.5)) +
  scale_fill_manual('legend', values = c('Ian' = '#C0C0C0', 'Jack' = '#404040', 'Conor' = '#808080', 'Nik' = '#1cd155'))
grid::grid.raster(logo3, 
                    x = 0.93, 
                    y = 0.9, 
                    just = c('left', 'bottom'), 
                    width = unit(0.35, 'inches'))
grid::grid.raster(logo2, 
                    x = 0.78, 
                    y = 0.02, 
                    just = c('left', 'bottom'), 
                    width = unit(1.5, 'inches'))
```
- Nik contributes most to the loudness of the playlist. Loudness was one of the top 3 audio features Nik answered that he considered when adding a song to the playlist!

```{r}
# average energy by user - might be related to loudness?
data %>% group_by(USER) %>% summarise(avgEnergy = mean(ENERGY), .groups='drop') %>% arrange(-avgEnergy)
```

# Popularity
```{r}
# reordering factors for plotting purposes
data$USER <- factor(data$USER, levels = c('Jack', 'Conor', 'Nik', 'Ian'))
data %>% 
  mutate(Popular = ifelse(POPULARITY > 20, 'Popular', 'Unpopular')) %>% 
  ggplot(aes(x=POPULARITY, fill=Popular)) +
  geom_histogram(position = 'identity', binwidth = 20, alpha=0.8) +
  scale_fill_manual(values=c('#1cd155', 'darkgray')) +
  facet_grid(~USER) +
  labs(title = 'Popularity Contribution by User',
       subtitle = 'Hungry Guys Radio on Spotify',
       caption = 'While Jack contributes the most songs overall, he has the lowest percentage of songs 
that are popular at 78%. Ian has the highest at 86%
       Data acquired via "Spotipy" API',
       x = '',
       y = 'Songs Added', 
       fill='') +
  theme_grey() + 
  theme(plot.title = element_text(color = 'black', face = 'bold', size = 16, hjust = 0.5),
        plot.caption = element_text(color = "black", face = 'italic', hjust = 0),
        plot.subtitle = element_text(color = "black", face = 'italic', hjust = 0.5))
grid::grid.raster(logo3, 
                    x = 0.93, 
                    y = 0.9, 
                    just = c('left', 'bottom'), 
                    width = unit(0.35, 'inches'))
grid::grid.raster(logo2, 
                    x = 0.785, 
                    y = 0.0, 
                    just = c('left', 'bottom'), 
                    width = unit(1.5, 'inches'))

```

```{r}
# breakdown of songs added
data %>% 
  select(USER, POPULARITY) %>% 
  mutate(Popular = ifelse(POPULARITY > 20, 1, 0)) %>% 
  group_by(USER) %>% 
  summarise(`Songs Added` = n(),
            `Popular Songs Added` = sum(Popular),
            `Average Popularity` = round(mean(POPULARITY), 2),
            `Median Popularity` = median(POPULARITY),
            `Percentage Popular Songs` = round((`Popular Songs Added` / `Songs Added`)*100, 2), .groups='drop') %>% 
  arrange(-`Percentage Popular Songs`)
```
- Jack has added the most popular (and unpopular) songs, however, he's added the most songs in general. Ian has the highest average popularity, however, he's only added 7 songs. Ian has the higher percentage of songs added that are popular at 85.71% and Jack has the lowest at 77.99%. Ian also indicated that he "maybe / sometimes" avoids adding popular songs to the playlist. Nik, however, indicated that he does avoid adding popular songs which shows in the data! Jack and Conor also answered "maybe / sometimes".

```{r}
# plotting popular and unpopular songs (relative to the playlist)
data %>% 
  mutate(popular = as.factor(ifelse(POPULARITY > 20, 'Popular', 'Unpopular'))) %>% 
  ggplot(aes(x=popular, y=ENERGY, fill = popular)) +
  geom_boxplot(color='black', alpha=0.8) +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  scale_fill_manual(values=c('#1cd155', 'darkgray')) +
  labs(title = 'What makes a song popular?',
       subtitle = 'Hungry Guys Radio on Spotify',
       caption = 'Popular songs have higher energy levels than unpopular ones.
       Data acquired via "Spotipy" API',
       x = '',
       y = 'Energy Level') +
  theme_grey() +
  theme(plot.title = element_text(color = 'black', face = 'bold', size = 15, hjust = 0.5),
        plot.caption = element_text(color = "black", face = 'italic', hjust = 0),
        plot.subtitle = element_text(color = "black", face = 'italic', hjust = 0.5),
        strip.text = element_text(size=8),
        legend.position = 'none')
  grid::grid.raster(logo3, 
                    x = 0.93, 
                    y = 0.9, 
                    just = c('left', 'bottom'), 
                    width = unit(0.35, 'inches'))
  grid::grid.raster(logo2, 
                    x = 0.785, 
                    y = 0.0, 
                    just = c('left', 'bottom'), 
                    width = unit(1.5, 'inches'))

```

# Valence
```{r}
# reordering factors for plotting purposes
data$USER <- factor(data$USER, levels = c('Jack', 'Conor', 'Ian', 'Nik'))
data %>% 
  ggplot(aes(x=USER, y=VALENCE, fill = USER)) +
  geom_boxplot(color='black', alpha=0.8) +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  scale_fill_manual('legend', values = c('Ian' = '#808080', 'Jack' = '#1cd155', 'Conor' = '#404040', 'Nik' = '#C0C0C0')) +
  labs(title = 'Valence Contribution by User',
       subtitle = 'Hungry Guys Radio on Spotify',
       caption = 'Check out and follow the playlist by scanning this Spotify code!
       Data acquired via "Spotipy" API',
       x = '',
       y = 'Valence Level') +
  theme_grey() +
  theme(plot.title = element_text(color = 'black', face = 'bold', size = 15, hjust = 0.5),
        plot.caption = element_text(color = "black", face = 'italic', hjust = 0),
        plot.subtitle = element_text(color = "black", face = 'italic', hjust = 0.5),
        strip.text = element_text(size=8),
        legend.position = 'none')
  grid::grid.raster(logo3, 
                    x = 0.93, 
                    y = 0.9, 
                    just = c('left', 'bottom'), 
                    width = unit(0.35, 'inches'))
  grid::grid.raster(logo2, 
                    x = 0.785, 
                    y = 0.0, 
                    just = c('left', 'bottom'), 
                    width = unit(1.5, 'inches'))

```


# Time
```{r}
# convert to date
data$DATE_ADDED <- mdy(data$DATE_ADDED)
data$DATE_ADDED_MY <- format(as.Date(data$DATE_ADDED), '%m-%Y')

addedAt <- data %>% 
  select(DATE_ADDED) %>% 
  filter(DATE_ADDED > '2019-09-20') %>% # filtering out initial playlist population
  group_by(month(DATE_ADDED, label = TRUE, abbr = FALSE)) %>% 
  summarise('Songs Added' = n(), .groups='drop')

names(addedAt)[1] <- 'Month'

```

```{r}
# songs additions by month
addedAt %>% 
  ggplot(aes(x=Month, y=`Songs Added`, fill = Month)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual('legend', values = c('January' = '#808080', 'February' = '#1cd155', 'March' = '#808080', 'April' = '#808080', 'May' = '#808080', 'June' = '#1cd155', 
                                         'July' = '#808080', 'August' = '#808080', 'September' = '#1cd155', 'October' = '#808080', 'November' = '#808080', 'December' = '#808080')) +
  geom_hline(yintercept = 16.25, color = "#1cd155", linetype = "dashed") +
  scale_x_discrete(guide = guide_axis(angle = 35)) +
  labs(title = 'Songs Added by Month',
       subtitle = 'Hungry Guys Radio on Spotify',
       caption = 'Check out and follow the playlist by scanning this Spotify code!
       Data acquired via "Spotipy" API',
       x = '',
       y = 'Songs Added') +
  theme_grey() +
  theme(plot.title = element_text(color = 'black', face = 'bold', size = 15, hjust = 0.5),
        plot.caption = element_text(color = "black", face = 'italic', hjust = 0),
        plot.subtitle = element_text(color = "black", face = 'italic', hjust = 0.5),
        strip.text = element_text(size=8),
        legend.position = 'none')
  grid::grid.raster(logo3, 
                    x = 0.93, 
                    y = 0.9, 
                    just = c('left', 'bottom'), 
                    width = unit(0.35, 'inches'))
  grid::grid.raster(logo2, 
                    x = 0.785, 
                    y = 0.0, 
                    just = c('left', 'bottom'), 
                    width = unit(1.5, 'inches'))

```

## How does Hungry Guys Radio compare to Spotify's "Today's Top Hits" playlist?
```{r}
hits <- read_csv('/Users/nik.linnane/Downloads/Spotify/topHits.csv')
hitsLogo <- image_read('/Users/nik.linnane/Downloads/Spotify/TTH2.png')

# standardize columns names
names(hits) <- toupper(names(hits))
hits$DATE_ADDED <- mdy(hits$DATE_ADDED)
hits$DATE_ADDED_MY <- format(as.Date(hits$DATE_ADDED), '%m-%Y')

# bind together both playlist dataframes
data$PLAYLIST <- 'Hungry Guys Radio'
hits$PLAYLIST <- "Today's Top Hits"
all <- rbind(data, hits)
```

```{r}
# compare popularity of hungry guys and top hits
all %>% 
  ggplot(aes(x=PLAYLIST, y=POPULARITY, fill = PLAYLIST)) +
  geom_boxplot(color='black', alpha=0.8) +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  scale_fill_manual(values=c('#1cd155', 'darkgray')) +
  labs(title = "Popularity Scores Compared to Today's Top Hits",
       subtitle = 'Hungry Guys Radio on Spotify',
       caption = 'When comparing our popularity scores to those of a very popular playlist, we 
can see how we emphasize adding less well-known songs!',
       x = '',
       y = '') +
  theme_grey() +
  theme(plot.title = element_text(color = 'black', face = 'bold', size = 15, hjust = 0.5),
        plot.caption = element_text(color = "black", face = 'italic', hjust = 0),
        plot.subtitle = element_text(color = "black", face = 'italic', hjust = 0.5),
        strip.text = element_text(size=8),
        legend.position = 'none')
  grid::grid.raster(logo3, 
                    x = 0.93, 
                    y = 0.9, 
                    just = c('left', 'bottom'), 
                    width = unit(0.35, 'inches'))
```

- today's top hits is also a LOT more consolidated  

```{r}
# can we find any relaytionship between audio features and popularity with a basic regression?
modelData1 <- data %>% select(DANCEABILITY, ENERGY, KEY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, TEMPO, DURATION_MIN, ACOUSTICNESS, EXPLICIT, POPULARITY)
modelData2 <- hits %>% select(DANCEABILITY, ENERGY, KEY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, TEMPO, DURATION_MIN, ACOUSTICNESS, EXPLICIT, POPULARITY)
modelData <- rbind(modelData1, modelData1)

# linear model to further understand relationship between track features and popularity
model <- lm(POPULARITY ~ DANCEABILITY + ENERGY + KEY + LOUDNESS + SPEECHINESS + INSTRUMENTALNESS + LIVENESS + VALENCE + 
              TEMPO + DURATION_MIN + ACOUSTICNESS + EXPLICIT, data = modelData)
summary(model)
```
- no relationship. could be the premise of another project?

## Create Machine Learning Dataset
```{r}
# import playlists datasets
altGood <- read_csv('/Users/nik.linnane/Downloads/Spotify/itsAltGood.csv')
jazz <- read_csv('/Users/nik.linnane/Downloads/Spotify/allNewJazz.csv')
morgan <- read_csv('/Users/nik.linnane/Downloads/Spotify/morganFreeman.csv')
country <- read_csv('/Users/nik.linnane/Downloads/Spotify/hotCountry.csv')
notSing <- read_csv('/Users/nik.linnane/Downloads/Spotify/cantNotSing.csv')
newTestData <- read_csv('/Users/nik.linnane/Downloads/Spotify/testingData.csv')

# standardize columns names
names(altGood) <- toupper(names(altGood))
altGood$DATE_ADDED <- mdy(altGood$DATE_ADDED)
altGood$DATE_ADDED_MY <- format(as.Date(altGood$DATE_ADDED), '%m-%Y')

names(jazz) <- toupper(names(jazz))
jazz$DATE_ADDED <- mdy(jazz$DATE_ADDED)
jazz$DATE_ADDED_MY <- format(as.Date(jazz$DATE_ADDED), '%m-%Y')

names(morgan) <- toupper(names(morgan))
morgan$DATE_ADDED <- mdy(morgan$DATE_ADDED)
morgan$DATE_ADDED_MY <- format(as.Date(morgan$DATE_ADDED), '%m-%Y')

names(country) <- toupper(names(country))
country$DATE_ADDED <- mdy(country$DATE_ADDED)
country$DATE_ADDED_MY <- format(as.Date(country$DATE_ADDED), '%m-%Y')

names(notSing) <- toupper(names(notSing))
notSing$DATE_ADDED <- mdy(notSing$DATE_ADDED)
notSing$DATE_ADDED_MY <- format(as.Date(notSing$DATE_ADDED), '%m-%Y')

names(newTestData) <- toupper(names(newTestData))
newTestData$DATE_ADDED <- mdy(newTestData$DATE_ADDED)
newTestData$DATE_ADDED_MY <- format(as.Date(newTestData$DATE_ADDED), '%m-%Y')


# combine
data2 <- data %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)
hits2 <- hits %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)
altGood2 <- altGood %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)
jazz2 <- jazz %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)
morgan2 <- morgan %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)
country2 <- country %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)

notSing2 <- notSing %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)

newTestData2 <- newTestData %>% select(ARTIST, ALBUM, TRACK_NAME, DANCEABILITY, ENERGY, LOUDNESS, SPEECHINESS, INSTRUMENTALNESS, LIVENESS, VALENCE, 
                         ACOUSTICNESS, EXPLICIT, POPULARITY, TEMPO, DURATION_MIN, PLAYLIST)

MLdata <- rbind(data2, hits2, altGood2, jazz2, morgan2, country2, notSing2)

# create labels
MLdata <- MLdata %>% mutate(HGR = ifelse(PLAYLIST == 'Hungry Guys Radio', 1, 0))
```

```{r}
# export
write_csv(MLdata, '/Users/nik.linnane/Downloads/Spotify/mlData2.csv')
```

```{r}
# export
write_csv(newTestData2, '/Users/nik.linnane/Downloads/Spotify/testingData.csv')
```


































